---
title: "Hands-on session: GRANAR – Generator of root anatomies in R"
author: "Adrien Heymans"
date: "2025-09-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(granar)
```

## introduction

In this session, we will learn how to generate and manipulate root anatomies with GRANAR.
We will:

- load and modify parameter files
- generate different types of root anatomies
- visualize anatomical structures with different options
- export data for use in MECHA simulations

### Making a first anatomy

We start by loading a simple monocot root parameter file, modifying the xylem features, and visualizing the result.

```{r}
#### LOAD - MODIFY - RUN - VISU #### 

# Load generic parameter file for GRANAR
params <- read_param_xml(path = "./www/root_monocot_simpl.xml")
# Xylem parameter
params$value[params$type == "max_size" & params$name == "xylem"] = 0.05
params$value[params$type == "n_files" & params$name == "xylem"] = 5
# RUN
root_sim <- create_anatomy(parameters = params)
# VISU
plot_anatomy(root_sim)
# Export anatomy
write_anatomy_xml(root_sim, "../MECHA/cellsetdata/current_root.xml")
aer_in_geom_xml(root_sim, "../MECHA/Projects/granar/in/Maize_Geometry.xml")
```

## Exploring complexity

Explore more detailed anatomies using a parameter set with additional tissues and apoplastic barriers.
We invite you to dive in the '.xml' file

### Sensitivity analysis per Teams!

```{r}

# Load generic parameter file for GRANAR
params <- read_param_xml(path = "./www/root_monocot_simpl.xml")
# Xylem parameter
params$value[params$type == "" & params$name == ""] = 
# RUN
root_sim <- create_anatomy(parameters = params)
# VISU
plot_anatomy(root_sim)
# Export anatomy
write_anatomy_xml(root_sim, "../MECHA/cellsetdata/current_root.xml")
aer_in_geom_xml(root_sim, "../MECHA/Projects/granar/in/Maize_Geometry.xml")
file.copy("../MECHA/Projects/granar/in/Maize_Geometry_aer.xml",
          "../MECHA/Projects/granar/in/Maize_Geometry_aer_sensitivity.xml" , overwrite = T)

```

```{r}
params_complx <- read_param_xml(path = "./www/root_monocot.xml")
sim_complx = create_anatomy(parameters = params_complx)
# To visualize the anatomy and the scenario that are going to be tested.
# you can use the plot_anatomy function.
plot_anatomy(sim_complx, col = "segment", apo_bar = 1)
plot_anatomy(sim_complx, col = "segment", apo_bar = 2)
plot_anatomy(sim_complx, col = "segment", apo_bar = 3)
```

GRANAR is aslo able to run secondary growth pattern. 

```{r}
params_complx <- read_param_xml(path = "./www/root_dicot_secGrowth.xml")
sim_complx = create_anatomy(parameters = params_complx)
# To visualize the anatomy and the scenario that are going to be tested.
# you can use the plot_anatomy function.

plot_anatomy(sim_complx, col = "dist")

```


## Make your own!

```{r}
# Change the parameter directly in the .xml or with multiple 
# my_params$value[my_params$type == "" & my_params$name == ""] = 
my_params <- read_param_xml(path = "./www/root_monocot_simpl.xml") # "./www/my_root.xml"
my_sim = create_anatomy(parameters = my_params)

plot_anatomy(my_sim, col = "area")

write_anatomy_xml(my_sim, "../MECHA/cellsetdata/test_root.xml")

aer_in_geom_xml(my_sim, "../MECHA/Projects/granar/in/Maize_Geometry.xml")
file.copy("../MECHA/Projects/granar/in/Maize_Geometry_aer.xml",
          "../MECHA/Projects/granar/in/Maize_Geometry_aer_test.xml" , overwrite = T)

```



## Comparing root types

Different root types can be generated from their respective parameter sets. Here we compare lateral, seminal, and nodal roots of a maize plant.

```{r}
# load the parameter sets
params_lat <- read_param_xml(path = "./www/lateral_root.xml")
params_sem <- read_param_xml(path = "./www/seminal_root.xml")
params_nod <- read_param_xml(path = "./www/nodal_root.xml")

# generation of the anatomies
sim_lat = create_anatomy(parameters = params_lat)
sim_sem = create_anatomy(parameters = params_sem)
sim_nod = create_anatomy(parameters = params_nod)

# check of the anatomies
plot_anatomy(sim_lat, col = "segment", apo_bar = 2)
plot_anatomy(sim_sem, col = "segment", apo_bar = 2)
plot_anatomy(sim_nod, col = "segment", apo_bar = 2)


```


## Exporting data for other functional models

The simulated anatomies can be saved in XML format for later use with MECHA.
We will export both cell set data and geometrical information (with aerenchyma).

```{r}
# save the nodes data 
write_anatomy_xml(sim_lat, "../MECHA/cellsetdata/root_lat.xml")
write_anatomy_xml(sim_sem, "../MECHA/cellsetdata/root_sem.xml")
write_anatomy_xml(sim_nod, "../MECHA/cellsetdata/root_nod.xml")

# save the geometrical information - aerenchyma
aer_in_geom_xml(sim_lat, "../MECHA/Projects/granar/in/Maize_Geometry.xml")
file.copy("../MECHA/Projects/granar/in/Maize_Geometry_aer.xml",
          "../MECHA/Projects/granar/in/Maize_Geometry_aer_lat.xml" , overwrite = T)
aer_in_geom_xml(sim_sem, "../MECHA/Projects/granar/in/Maize_Geometry.xml")
file.copy("../MECHA/Projects/granar/in/Maize_Geometry_aer.xml",
          "../MECHA/Projects/granar/in/Maize_Geometry_aer_sem.xml" , overwrite = T)
aer_in_geom_xml(sim_nod, "../MECHA/Projects/granar/in/Maize_Geometry.xml")
file.copy("../MECHA/Projects/granar/in/Maize_Geometry_aer.xml",
          "../MECHA/Projects/granar/in/Maize_Geometry_aer_nod.xml" , overwrite = T)
```



## recap:

- You learned how to load and modify parameter files in GRANAR
- You generated and visualized different root anatomies
- You explored anatomical complexity with apoplastic barriers
- You exported anatomies for further simulations with MECHA
